---
title: "Base Pricing Analysis and Price Elasticity Estimation"
author: "Günter J. Hitsch"
date: "January 8, 2018"
output: pdf_document
urlcolor: blue
graphics: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE, eval = FALSE,
                      fig.width = 4.5, fig.height = 3, fig.align = "right")

stargazer_type = "text"      # "text" for running in Notebook mode, "latex" when creating a pdf document.
```



## Overview

The goal is to conduct a base pricing analysis. We estimate brand-level demand using scanner data, and then we make profitability predictions corresponding to specific base price changes. We estimate log-linear demand models that use (log) prices and promotions as inputs, and predict log quantities, `log(1+Q)`. The models predict the demand for a focal brand, and we control for (log) prices and promotions of three competitors. Obviously, this approach generalizes to an arbitrarily large number of competing products as long as the sample size is large enough.

Our focus is on the two top brands in the liquid laundry detergent category, *Tide* and *Gain*. Both are Procter & Gamble brands. The two closest competitors are *Arm & Hammer* and *Purex*.



\bigskip

## Packages

Make sure to install three packages that we have not used before: lfe, knitr, and stargazer.

```{r}
library(bit64)
library(data.table)
library(lfe)
library(knitr)
library(stargazer)
library(ggplot2)
```



\newpage

## Data overview

The data source is an extract from the Nielsen RMS retail scanner data set. The data set captures weekly price and quantity data for all products (UPC's) sold in the stores of a large number of U.S. retail chains. The Kilts data do not include all retailers (for example, Walmart is not part of the data), and the identity of the retailers is not revealed. However, we know if a store belongs to the same retail chain.


#### Brand data

The data.table `brands` in `Brands.RData` includes brand information for the top five brands in three categories (product modules):

```{}
1036   FRUIT JUICE - LEMON/LIME
1040   FRUIT JUICE - ORANGE - OTHER CONTAINER
7012   DETERGENTS - HEAVY DUTY - LIQUID
```

The data include the brand code, brand description, and total revenue calculated across all observations. The top five brands were selected based on total brand revenue.

We will focus on the liquid laundry detergent category with corresponding `product_module_code` 7012. 


#### Store data

Inspect the table `stores` in the file `Stores.RData`. `store_code_uc` identifies each retail stores. For some (but not all) stores we know the corresponding `retailer_code` that identifies the chain (banner) that the store belongs to. The data include the Scantrack (SMM) market code and the Scantrack market description. Scantrack markets correspond to large metropolitan market areas such as *Chicago* or *Raleigh-Durham* (see the data manual for a map of the Scantrack markets). The three-digit ZIP code of each store is also included.


#### Movement data

The movement data (`move`) are in files of the form `brand_move_<module code>.RData`. The data are at the brand/store/week level and include prices and quantities (`units`). The data are aggregates of all UPC's that share the same brand name. Brand prices are measured as the weighted average over all store/week UPC prices in equivalent units, and quantities represent total product volume measured in equivalent units such as ounces. In the liquid laundry detergent category (module 7012), prices represent dollars per ounce and units are total product volume in ounces per store/week. The aggregation weights are based on total store-level UPC revenue across all weeks, and hence the aggregation weights are constant within each store. The movement data also include a promotion indicator (`promo_dummy`), a logical `TRUE/FALSE` variable.

The `week_end` variable date is the last day of a Nielsen week, which always starts on a Sunday and ends on a Saturday. Note that prices may change during the period, and hence even the UPC-level price may be an average over more than one posted price. The sample includes data for the 2010-2013 period.

Please consult the official Kilts Center Retail Scanner Dataset Manual for all details.



\newpage

## Prepare the data for the demand analysis

We first load the brand and store meta data.

```{r}
load("./Data/Brands.RData")
load("./Data/Stores.RData")
```


#### Select the category and brands

*Choose the laundry detergent category (module) and select the corresponding brand-level meta data from the data table `brands`. Then sort (order) the brand data corresponding to total brand revenue, and select the **top four brands** (ranked by revenue). You may (completely optionally) verify that the predictions of the analysis remain robust if you include all top five brands.*

```{r}
selected_module = 7012                 # Laundry detergent
```

\medskip

*Let's assign each brand a new name using a new variable, `brand_name`, and give the four brands simple names such as `Tide`, `Gain`, `ArmHammer`, and `Purex`. These simplified brand names will make our code and the estimation output more readable.*

Note that we will add the brand names to the quantity, price, and promotion variables. In R, `price_ArmHammer` (as well as `price_Arm_Hammer`) are legal variable names, but `price_Arm&Hammer` and `price_Arm & Hammer` are not, and hence I do not suggest the brand names `Arm&Hammer` or `Arm & Hammer`.


#### Prepare the movement data

*Load the movement data, and---for better readability---change the variable names from `units` to `quantity` and from `promo_dummy` to `promotion`. Change the data type of the `promotion` variable from `logical` to `numeric` using the `as.numeric` function. Finally, merge the new `brand_name` variable with the movement table.*


#### Remove outliers

Most data contain some "flaws" or outliers. Here is an easy way of removing such outliers:

First, we create a function that flags all observations in a vector `x`, for example a price series, as outliers if the ratio between a value and the median value among all `x` observations is below or above a threshold.

```{r}
isOutlier <- function(x, threshold_bottom, threshold_top) {
   is_outlier = rep(FALSE, times = length(x))
   median_x   = median(x, na.rm = TRUE)
   is_outlier[x/median_x < threshold_bottom | x/median_x > threshold_top] = TRUE
   return(is_outlier)
}
```

*Now run this function on the price data, separately for each brand and store. Then tabulate the number of outliers, and remove the corresponding observations from the data set.*

I recommend to use a lower threshold (`threshold_bottom`) value of 0.35 and an upper threshold (`threshold_top`) of 2.5.


#### Reshape the movement data from long to wide format

To prepare the data for the regression analysis, we need to **reshape the data from long to wide format** using **`dcast`**.

All the details on casting and the reverse operation, melting from wide to long format using `melt`, are explained in the data.table html vignettes: <https://rawgit.com/wiki/Rdatatable/data.table/vignettes/datatable-reshape.html>.

Let's be specific about the structure of the data that we need to use to estimate a demand model. We would like to obtain a table with observations, characterized by a combination of store id (`store_code_uc`) and week (`week_end`) in rows, and information on quantities, prices, and promotions in columns. Quantities, prices, and promotions are brand-specific. Hence, the structure of the wide-format data.table that we want to create is

`store_code_uc + week_end ~ brand_name`,

and the brand-specific variables that we would like have in the columns are

`value.var = c("quantity", "price", "promotion")`.

*Let's use `dcast` to obtain the corresponding data*:

```{r}
move = dcast(move, store_code_uc + week_end ~ brand_name, 
             value.var = c("quantity", "price", "promotion"))
head(move)
```


#### Merge store information with the movement data

*Now merge the movement data with the store meta data, in particular with the retailer code, the Scantrack (SMM) market code, and the Scantrack market description. But only with the store meta data where we have a valid retailer code. Hence, we need to remove store data if the retailer code is missing (`NA`). Use the `is.na` function to check if `retailer_code` is `NA` or not.*


#### Create time variables or trends

*Extract the year and month from the week (`week_end`) variable in the movement data (use the functions `year` and `month`) and create a time trend (1, 2, 3, ...) such that 1 corresponds to a week in the first month in the data, 13 corresponds to a week in the 13th month in the data, etc.*


#### Remove missing values

Finally, *retain only complete cases*, i.e. rows without missing values:

```{r}
move = move[complete.cases(move)]
```



\newpage

## Data inspection

#### Observations and geographic coverage

*First, document the number of observations and the number of unique stores in the data.*

*Second, we assesss if the included stores have broad geographic coverage. We hence create a summary table that records the number of observations for each separate Scantrack market:*

```{r}
market_coverage = move[, .(n_obs = .N), by = SMM_description]
```

Note the use of the data.table internal `.N`: `.N` is the number of observations, either in the whole data table, or---as in this case---the number of observations within each group defined by the `by =` statement.

\medskip

A convenient way to print a table is provided by the **`kable`** function that is included in the `knitr` package. Please consult the documentation for `kable` to see all options. Particularly useful are the options `col.names`, which is used below, and `digits`, which allows you to set the number of digits to the right of the decimal point.

*Now use `kable` to document the number of observations within each Scantrack market.*

```{r}
kable(market_coverage, col.names = c("Scantrack market", "No. obs."))
```


#### Price variation

Before estimating the demand models we would like to understand the degree of price variation in the data. Comment on why this is important for a regression analysis such as demand estimation!

We will predict demand for Tide and Gain. For each of these two brands separately, we would like to visualize the overall degree of price variation across observations, and also the variation in relative prices with respect to the competing brands.

- *To visualize the (own) price variation, normalize the prices of Tide and Gain with respect to the average of these prices, and show the histogram of normalized prices.*

- *To visualize relative prices, calculate the ratio of Tide and Gain prices with respect to the three competiting brands, and show the histogram of relative prices.*

Note: To avoid that the scale of a graph is distorted by a few outliers, use the `limits` option in `scale_x_continuous` (see the ggplot 2 introduction). This also helps to make the graphs comparable with each other.


#### Summary of data inspection

*Discuss the data description, including sample size, geographic coverage, and the results on own and relative price variation.*



\newpage

## Estimation

Now we are ready to estimate demand models for Tide and Gain.

We want to estimate a sequence of models with an increasing number of controls and compare the stability of the key results across these models. In all models the output is `log(1+quantity_<brand name>)`.

\bigskip

To keep things simple, we will initially estimate demand for Tide only.

Let's start with the following models:

1. log of own price as only input
2. Add store fixed effects
3. Add a time trend---maybe linear, or a polynomial with higher-order terms
4. Instead of a time trend add fixed effects for each month (more precisely: for each year/month combination)

*Estimate the models using the `felm` function from the lfe package (consult the corresponding notes on Canvas). Store the outputs in some appropriately named variables (objects).*

\bigskip

**Hint**: Recall that it is perfectly legitimate in R to write model formulas such as

```{}
log(1+quantity_<brand name>) ~ log(price_<brand name>)
```

Hence, there is no need to create new variables such as the logarithm of own price, etc., before estimating a demand model.

\bigskip

You can display the regression coefficients using the `summary` function, which is standard. As a much more elegant solution, however, I recommend to use the stargazer package to produce nicely formatted output. Here is an easy example of how to use stargazer (note that the `fit` objects are the regression outputs, adjust the names if necessary):

```{r}
stargazer(fit_base, fit_store_FE, fit_trend, fit_month_FE, 
          type  = "text",
          column.labels  = c("Base", "Store FE", "Trend", "Store + year/month FE"),
          dep.var.labels.include = FALSE)
```

As you see, stargazer automatically displays the estimates for the same variable in the same row. This vastly improves the readability and comparability of the estimates across models.

To learn more, consult the stargazer documentation or <http://jakeruss.com/cheatsheets/stargazer.html>.

\bigskip

Before moving on, you may want to remove the regression output objects that are no longer used, because they take up much space in memory:

```{r}
rm(fit_base, fit_store_FE, fit_trend)
```


#### Controlling for competitor prices

*Now add the competitor prices to the demand model.*


#### Controlling for promotions

*Now add the promotions dummies, first just for Tide, then for all brands. Compare the results. Did controlling for promotions change the own price elasticity estimate in an expected manner?*


\bigskip

*Summarize and comment on the estimation results. Was it necessary to control for store fixed effects, time trends/fixed effects, as well as competitor prices and promotions? What do we learn from the magnitudes of the own and cross-price elasticities?*

\bigskip

We will use the final model including all variables (I called it `fit_promo_comp`) as our preferred model. Also, we *save the model output object in a named file in the folder `Results`*. Make sure to create this folder in the same directory where the R Markdown source file is located.

Before saving `fit_promo_comp` we rename the regression output object to make it distinguishable from the regression output for Gain.

```{r}
fit_Tide = fit_promo_comp 
save(fit_Tide, file = "./Results/fit_Tide.RData")
```


\bigskip

**Warning**: The estimation output object is typically large, because R includes all the original data in the object. To use stargazer, the whole estimation output object is needed. Other summaries can be obtained using only the result of the `summary` function:

```{r}
summary_promo_comp = summary(fit_promo_comp)
```

This summary object is of considerably smaller size than the original output object.

```{r}
print(object.size(fit_promo_comp), units = "Mb")
print(object.size(summary_promo_comp), units = "Mb")
```


#### Demand model for Gain

*Now repeat the steps to estimate demand for Gain, and store the final estimation output.*

*Briefly comment on the estimates, as you did before with Tide.*



\newpage

## Profitability analysis

The goal is to fine-tune prices jointly for Tide and Gain. We hence use the estimates of the preferred demand models and evaluate the product-line profits when we change the prices of the two brands.

\bigskip

To predict profits we need to predict demand using the regression output. However, `felm` does not have an associated predict function, and hence I created my own (see my notes on the lfe package). Make sure to download and source the `predict.felm.R` script.

```{r}
source("./predict.felm.R")
```

\bigskip

To predict profits, let's only retain data for one year, 2013:

```{r}
move_predict = move[year == 2013]
```

\bigskip

Although we have excellent demand data, we do not know the production costs of the brands (this is confidential information). We can infer the cost making an informed assumption on retail margins and the gross margin of the brand.  

```{r}
gross_margin  = 0.35
retail_margin = 0.18

cost_Tide = (1-gross_margin)*(1-retail_margin)*mean(move_predict$price_Tide)
cost_Gain = (1-gross_margin)*(1-retail_margin)*mean(move_predict$price_Gain)
```

As prices are measured in dollars per ounce, these marginal costs are also per ounce.

\bigskip

Now create a vector indicating the percentage price changes that we consider within an acceptable range, up to +/- five percent.

```{r}
percentage_delta = seq(-0.05, 0.05, 0.025)    # Identical to = c(-0.5, -0.025, 0.0, 0.025, 0.05)
```

\bigskip

We will consider all possible combinations of price changes for Tide and Gain. This can be easily achieved by creating a data table with the possible combinations in rows (please look at the documentation for the `rep` function):

```{r}
L = length(percentage_delta)
profit_DT = data.table(delta_Tide = rep(percentage_delta, each = L),
                       delta_Gain = rep(percentage_delta, times = L),
                       profit     = rep(0, times = L*L))
```

Inspect the resulting table. The `profit` column will allow us to store the predicted profits.

\bigskip

Now we are ready to iterate over each row in `profit_DT` and evaluate the total product-line profits of Tide and Gain for the corresponding percentage price changes. You can perform this iteration with a simple for-loop:

```{r, eval = FALSE}
for (i in 1:nrow(profit_DT)) {
   # Perform profit calculations for the price changes indicated in row i of the profit_DT table
}
```

\medskip

Some hints:

- Before you start the loop, store the original price levels of Tide and Gain.
- Update the price columns in `move_predict` and then predict demand.
- Calculate total profits at the new price levels for both brands and then store the total profit from Tide and Gain in `profit_DT`.

\medskip

Show a table of profits in levels and in ratios relative to the baseline profit at current price levels, in order to assess the percent profit differences resulting from the contemplated price changes.

\bigskip

*Discuss the profitability predictions and how prices should be changed, if at all. How do you reconcile the recommended price changes with the own-price elasticity estimates?*




\newpage

## Appendix

#### Professionally formatted output in stargazer (completely optional!)

If you knit your R Markdown source to a pdf file, and if you would like to create very professional-looking tables, you can instruct stargazer to output the table in LaTeX. Simply set `type = "latex"` instead of `type = "text"`.

```{r, results = "asis"}
stargazer(fit_base, fit_store_FE, fit_trend, fit_month_FE, 
          type  = "latex",
          column.labels  = c("Base", "Store FE", "Trend", "Store + year/month FE"),
          dep.var.labels.include = FALSE,
          header = FALSE)
```

Note the added option `header = FALSE`.

\bigskip

Furthermore and very **importantly**, in the R code chunk header above the stargazer call, you need to include the option `results = "asis"`.

\bigskip

Typically you will use a workflow where you first develop your R Markdown file in Notebook mode. During this phase producing LaTeX output is not desirable and yields unreadable results. Once the script is finished, you compile the R Markdown source to produce the final output. Instead of manually changing the output type from text to LaTeX you can use the following trick.

Create a flag at the top of the R Markdown source, such as

```{r}
stargazer_type = "text"      # "text" for running in Notebook mode, "latex" when creating a pdf document.
```

You can hide this flag from the final output by adding the option "include=FALSE" in the R code chunk header.

Ensure that each stargazer call uses this flag to set the desired output type:

```{r, results = "asis"}
stargazer(fit_base, fit_store_FE, fit_trend, fit_month_FE, 
          type  = stargazer_type,
          column.labels  = c("Base", "Store FE", "Trend", "Store + year/month FE"),
          dep.var.labels.include = FALSE,
          header = FALSE)
```

Now you only need to change the flag when you produce the final output from `= "text"` to `= "latex"` to change the output type for all tables.


