---
title: "Private Label Demand"
author: "Jack Gang, Lakshmi Jampana, Jennifer Lopez, Tommy Wu"
date: "1/31/2017"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
header-includes: \usepackage{color}
graphics: yes
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, message = FALSE,
                      fig.width = 10, fig.height = 10, fig.align = "center")
```

In this assignment, we will answer the question of whether private label demand is sensitive to household income and wealth. We will study this question using data from the Nielsen Homescan panel, and we will also use publicly available local home value data from Zillow as a measure of wealth. The overall data analysis strategy is as follows:

1. Construct household-level private label shares for all months between 2004-2014
2. Use household-level income, employment, and other information contained in the Nielsen data
3. Use local, 5-digit ZIP code Zillow home value data as a measure of local housing wealth
4. Summarize the data
5. Provide regression results of private label shares on income, employment, wealth (Zillow home value index), and other controls

In the first part of the assignment we will construct the private label share data and save them to file. In the second part we will finish the data preparation and then run the statistical analysis.

## Zillow data construction

We first retrieved the Zillow data (in csv format) from the Zillow website:


```{r}
library(bit64)
library(data.table)
library(lubridate)
```

```{r}
# read in Zillow data
zillow_DT = fread("http://files.zillowstatic.com/research/public/Zip/Zip_Zhvi_AllHomes.csv")
names(zillow_DT)[1:20]
```

\newpage

Next, we `melt`ed the table from wide to long format:

```{r}
# melt zillow_DT
zillow_DT = melt(zillow_DT, id.vars = names(zillow_DT)[1:7], measure.vars = 
                   names(zillow_DT)[8:ncol(zillow_DT)], variable.name = "date",
                 value.name = "zillow_index")
head(zillow_DT)
```

Before saving the data, we wanted to clean it. First, we converted `date` to a `month_year` R date variable:

```{r}
# convert date to month_year
zillow_DT[, date := as.character(date)]
zillow_DT[, month_year := ymd(date, truncated = 2)]
```

We then created a `zip_code` variable from `RegionName`:

```{r}
zillow_DT[, zip_code := as.integer(RegionName)]
```

Finally, we removed some unnecessary variables, set the preferred column order, and keyed the data:

```{r}
zillow_DT[, c("date", "RegionID", "RegionName") := NULL]
setcolorder(zillow_DT, c("zip_code", "month_year", "zillow_index","City",
                         "State", "Metro", "CountyName", "SizeRank"))
setkey(zillow_DT, zip_code, month_year)
```

After cleaning the data, we saved it to a file:

```{r, eval = FALSE}
save(zillow_DT, file = "/classes/37105/group7/Assignment 3/Data/Zillow-Data.RData")
```

\newpage

## Household-level private label share creation

### Tagging private label products

We used the same Nielsen Homescan household panel data for the analysis as we did in the first assignment. The first thing we did was analyze the evolution of private label shares. We loaded the product data and added a flag called `is_PL` that equals `TRUE` if a product is a private label product:

```{r}
# load the product data
load("/classes/37105/main/Assignment-3/Products.RData")

# add a flag to mark if a product is a private label
products[, is_PL := ifelse(like(brand_descr, "CTL BR"), TRUE, FALSE)]

# look at table of all the matches for private label
table(products[products$is_PL]$brand_descr)
```

### Iteration and data reduction strategy

Since the purchase data are at the year level, we iterated over all the files to combine the calculations in one table. For each household, we calculated private label shares at the year/month level:

```{r, eval = FALSE}
shares_DT = data.table()

# iterate over all purchase files
for (yr in 2004:2014)
{
  # load purchase data
  load(paste0("/classes/37105/main/Assignment-3/purchases_", yr, ".RData"))
  
  # merge is_PL flag, department code, and product module code
  purchases = merge(purchases, 
                    products[, .(upc, upc_ver_uc, is_PL, department_code, 
                                 product_module_code)])
  
  # drop some departments and product modules from the data set
  department_DT = products[, head(.SD, 1), by = department_code,
                     .SDcols = c("department_descr")]
  department_DT[order(department_code)]
  
  purchases = purchases[purchases$department_code != 9] # general merchandise
  purchases = purchases[purchases$department_code != 99] # magnet data
  purchases = purchases[purchases$product_module_code >= 1000] # magnet data 
  purchases = purchases[!is.na(purchases$department_code)] # unknown department code
  
  # create a month_year date variable
  purchases[, month_year := floor_date(purchases$purchase_date, "month")]
  
  # create total dollar spending at the household/month/year level, separately 
  # for private label and other (national brand) products
  purchases[, no_households := length(unique(household_code)), by = month_year]
  purchases_total = purchases[, .(totalSpend = sum(total_price_paid - coupon_value)),
                           keyby = .(is_PL, household_code, month_year)]
  purchases_total = dcast(purchases_total, household_code + month_year ~ is_PL, 
               value.var = c("totalSpend"))
  
  # convert to percentage shares
  purchases_total[, perc_share := purchases_total$"TRUE"/
                    (purchases_total$"TRUE" + purchases_total$"FALSE")]
  
  # keep data on household/month/year/private label shares only
  purchases_total[, c("TRUE","FALSE"):=NULL]
  
  # insert the final data for the given year into a list
  shares_DT = rbind(shares_DT, purchases_total)
}

# remove missing values due to 0 household spending
nrow(shares_DT)
shares_DT = shares_DT[complete.cases(shares_DT)]
nrow(shares_DT)

# key the data at the household/month/year level save to file
setkeyv(shares_DT, c("household_code", "month_year"))
save(shares_DT, file = "/classes/37105/group7/Assignment 3/Data/shares-Data.RData")

```

## Household data preparation

Before we began with the main analysis, we first examined the household data in the `panelists.RData` file.

```{r}
library(psych)
library(lfe)
library(ggplot2)
library(knitr)
library(stargazer)

# load and examine panelist data
load("/classes/37105/main/Assignment-3/panelists.RData")
names(panelists)
```

\newpage

It is important to note that household income is represented as a factor with different levels representing an income range:

```{r}
# inspect household_income variable
is.factor(panelists$household_income)
levels(panelists$household_income)
```

Our results were easier to interpret if we converted the factor representation of income to a dollar measure as follows:

```{r}
# convert household_income to dollars
panelists[household_income == "-$5000", income := 2500]
panelists[household_income == "$5000-$7999", income := 6500]
panelists[household_income == "$8000-$9999", income := 9000]
panelists[household_income == "$10,000-$11,999", income := 11000]
panelists[household_income == "$12,000-$14,999", income := 13500]
panelists[household_income == "$15,000-$19,999", income := 17500]
panelists[household_income == "$20,000-$24,999", income := 22500]
panelists[household_income == "$25,000-$29,999", income := 27500]
panelists[household_income == "$30,000-$34,999", income := 32500]
panelists[household_income == "$35,000-$39,999", income := 37500]
panelists[household_income == "$40,000-$44,999", income := 42500]
panelists[household_income == "$45,000-$49,999", income := 47500]
panelists[household_income == "$50,000-$59,999", income := 55000]
panelists[household_income == "$60,000-$69,999", income := 65000]
panelists[household_income == "$70,000-$99,999", income := 85000]
panelists[household_income == "$100,000 + ", income := 112500]
panelists[household_income == "$100,000 - $124,999", income := 112500]
panelists[household_income == "$125,000 - $149,999", income := 132500]
panelists[household_income == "$150,000 - $199,999", income := 175000]
panelists[household_income == "$200,000 + ", income := 250000]
```

Because the detailed income levels above \$100,000 were only recorded for the 2006-2009 panel years, for consistency we replaced these values with \$112,500:

```{r}
# replace high incomes with $112,500
panelists[income >= 100000, income := 112500]
```

\newpage

Furthermore, Homescan's household income data represents ranges of total household income for the full year that is 2 years prior to the Panel Year. To correctly date the income data we needed to associate the income reported two years in the future with the current panel year:

```{r}
# make sure panelists table is correctly keyed
key(panelists)

# replace income for a given year with the lead of income two years ahead
panelists[, income := shift(income, n = 2, type = "lead"), by = household_code]
```

Some of the key demographic variables, in particular, age or birth year, employment, and education, are available both for a male head and a female head in each household. However, depending on the household composition, only one household head may be present, and more generally we want to simplify this information to make our results more easily interpretable.

A simple solution is to use information on employment, etc., for the male head by default, and use information for the female head if no male head is present in a household. The choice of the male head reflects that the incidence of the employment status `Not Employed for Pay` is higher for the female head than for the male
head in the data. We confirmed this:

```{r}
# incidence of employment status "Not Employed for Pay" by gender
nrow(panelists[panelists$male_head_employment == "Not Employed for Pay"])
nrow(panelists[panelists$female_head_employment == "Not Employed for Pay"])
```

Then we created new `age`, `unemployed`, and `education` variables:

```{r}
# female head flag
panelists[, female_head := male_head_age == "No Such Head"]

# age variable - converted from birth year character string
head(panelists$female_head_birth)
panelists[, age := male_head_birth]
panelists[female_head == TRUE, age := female_head_birth]
panelists[, age := panel_year - as.numeric(substr(age, 1, 4))]

# unemployed flag
panelists[, unemployed := male_head_employment == "Not Employed for Pay"]
panelists[female_head == TRUE, 
          unemployed := female_head_employment == "Not Employed for Pay"]

# education flag
panelists[, education := male_head_education]
panelists[female_head == TRUE, education := female_head_education]
```

\newpage

In addition, we created two more variables. `size` provides a numeric measure of the `household_size` of a panelist. `has_children` is a dummy indicator that equals 1 if children are present in the household:

```{r}
# size variable
levels(panelists$household_size)
panelists[household_size == "Single Member", size := 1]
panelists[household_size == "Two Members", size := 2]
panelists[household_size == "Three Members", size := 3]
panelists[household_size == "Four Members", size := 4]
panelists[household_size == "Five Members", size := 5]
panelists[household_size == "Six Members", size := 6]
panelists[household_size == "Seven Members", size := 7]
panelists[household_size == "Eight Members", size := 8]
panelists[household_size == "Nine+ Members", size := 9]

# has_children variable
levels(panelists$age_and_presence_of_children)
panelists[, has_children := TRUE]
panelists[age_and_presence_of_children == "No Children Under 18"]$has_children = FALSE
```

\newpage

## Merge demographic and home value data with the private label share data

We then merged the household data and Zillow home indices with our previously calculated private label share data.

### Merge demographic information

We created a common, unambiguous key between the share data and the panel data by extracting the year component from the date variable in the share data:

```{r}
# load in share data and create common year key
load("/classes/37105/group7/Assignment 3/Data/shares-Data.RData")
shares_DT[, panel_year := year(month_year)]
```

Next, we merged demographic variables, zip code, DMA code, and household projection factors with the share data:

```{r}
# merge panelist information with share data
shares_DT = merge(shares_DT, panelists[, .(household_code, panel_year, income,
                                           unemployed, education, age, size, 
                                           has_children, female_head, 
                                           marital_status, race, 
                                           hispanic_origin, panelist_zip_code, 
                                           dma_code, projection_factor)],
                  by=c("household_code","panel_year"))
```

### Merge Zillow data

Next, we merged the Zillow data with the shares data:

```{r}
# load in Zillow data and create common zip code key
load("/classes/37105/group7/Assignment 3/Data/Zillow-Data.RData")
setnames(shares_DT, "panelist_zip_code", "zip_code")

# merge Zillow data with share data
shares_DT = merge(shares_DT, zillow_DT[, .(zip_code, month_year, zillow_index,
                                           City, State, Metro, CountyName,
                                           SizeRank)], 
                  by=c("zip_code","month_year"), all.x = TRUE)

```

After all these merges, we set the key for the share data to return them to the original order:

```{r}
# key the share data to household/month/year level
setkeyv(shares_DT, c("household_code", "month_year"))
```

Finally, for better scaling, we represented the shares on a 0-100 percent scale:

```{r}
# represent shares on 0-100 percent scale:
shares_DT$perc_share = shares_DT$perc_share*100
```

\newpage

## Data description

### Distribution of private label shares across households

First, we provided an overview of the distribution of private label shares across households. To average away the randomness in the month-to-month shares, we created annual private label shares for each household based on a simple average (mean):

```{r}
# create annual private label shares for each household and plot histogram
shares_households = shares_DT[,.(mean_perc_share = mean(perc_share)), 
                              by = .(panel_year,household_code)]

```

### Provide summary statistics and a histogram of the distribution of the private label shares across households

Next, we summarized the private label shares data by looking at a histogram and summary statistics of the mean private label shares per household across all years. We can see that the majority (within 1 SD on both sides) of households have private label shares between 10-30%. A skew of 1.0 also indicates that the private label shares are skewed right (as confirmed by the histogram):

```{r}
# summarize PL share data by household (ignore year)
shares_summary = shares_households[, .(perc_share = mean(mean_perc_share)),
                           keyby = .(household_code)]

# summary statistics
describe(shares_summary$perc_share)

# show in histogram
ggplot(shares_summary, aes(perc_share)) + 
  geom_histogram() + scale_x_continuous("Private Label Shares - by household", 
                                        limits = c(0, 100)) +
  scale_y_continuous("Count", limits = c(0, 6000))
```

### Evolution of private label shares over time

To plot the evolution of private label shares over time, we provided a time-series graph of mean private label shares (across households) by month. We first calculated the average (across households) private label shares for each month/year using a weighted mean (with `projection_factor` as the weight):

```{r}
# calculate average private label shares for each month/year
shares_households = shares_DT[,.(mean_perc_share =
                                   weighted.mean(perc_share, projection_factor)), 
                              by = .(month_year)]

# time-series graph of evolution of private label shares
ggplot(shares_households, aes(month_year, mean_perc_share)) +
  annotate("rect", xmin = mdy("12-01-2007"), xmax = mdy("06-01-2009"), 
           ymin = -Inf, ymax = Inf, fill = "lightblue1", alpha = 0.4) +
  geom_line() +
  geom_point() +
  scale_x_date("Year", date_labels = "%Y", date_breaks = "1 years", minor_breaks = NULL)

```

\newpage

### Change in home values

To document the change in home values between June 2006 and June 2009 (the last month of the Great Recession), we calculated the percentage change in the `zillow_index` at the ZIP code level:

```{r}
# aggregate mean zillow_index data by month/year and zip code
zindex_monthzip = shares_DT[,.(mean_zillow_index = 
                                 weighted.mean(zillow_index, projection_factor)), 
                            by = .(month_year, zip_code)]
setkeyv(zindex_monthzip, c("zip_code", "month_year"))

# remove zip codes that don't have data for June 2006 or June 2009
zips = intersect(unique(zindex_monthzip[month_year == '2006-06-01']$zip_code), 
                 unique(zindex_monthzip[month_year == '2009-06-01']$zip_code))
zindex_monthzip2 = zindex_monthzip[zindex_monthzip$zip %in% zips]

# shift zillow_index by 36 and calculate the percentage change
zindex_monthzip2[, shifted_zillow_index := shift(mean_zillow_index, 36)]
zindex_monthzip2[, perc_change := shifted_zillow_index/mean_zillow_index]
perc_changes = zindex_monthzip2[month_year == '2009-06-01']
perc_changes = perc_changes[complete.cases(perc_changes)]

```

Then, we provided a histogram of the percentage home value changes (June 2009 versus June 2006):

```{r}
# plot histogram of percentage home value changes from June 2006 to June 2009
ggplot(perc_changes, aes(perc_change)) + 
  geom_histogram() + scale_x_continuous("% Home Value Change 6/2006 - 6/2009", 
                                        limits = c(0.2, 3.5)) +
  scale_y_continuous("Count", limits = c(0, 1000))
```

\newpage

## Main Analysis

We then estimated income and wealth effects on private label demand using regression analysis.

### Model specifications

We estimated 4 different models and clustered standard errors at the market/year level for each. This clustering ensured that we were not understating our standard errors, and we did indeed see that when using clustering, the standard error of our coefficients increased. 

First, the base model used the main variables, `log(income)`, the unemployed indicator, and the wealth measure (`log(zillow_index)`), as covariates in the regression. The dependent variable was the household-level private label share in each month. In this base model we pooled across all units in the panel:

```{r}
# base model
fit_base = felm(perc_share ~ log(income) + factor(unemployed) +
                  log(zillow_index), data = shares_DT)
```

We then added all the demographic controls in an attempt to control for heterogeneity across the units using observed information:

```{r}
# demographics model
fit_demo = felm(perc_share ~ log(income) + factor(unemployed) +
                  log(zillow_index) + factor(education) + age + size +
                  factor(has_children) + factor(female_head) + 
                  factor(marital_status) + factor(race) + 
                  factor(hispanic_origin), data = shares_DT)
```

Next, instead of demographics, we used household-level fixed effects instead to provide within-estimates of the main variables:

```{r}
# household fixed effects model
fit_household = felm(perc_share ~ log(income) + factor(unemployed) +
                       log(zillow_index) | household_code | 0 | dma_code + 
                       panel_year, data = shares_DT)
```

Lastly, we added different time controls, such as time trend, trend plus recession indicator (dummy), and month/year fixed effects:

```{r}
# create time trend and estimate time trend model
shares_DT[, time_trend := as.numeric(month_year)]
shares_DT[, time_trend := (time_trend - min(time_trend))/365]

# time trend model
fit_timetrend = felm(perc_share ~ log(income) + factor(unemployed) +
                       log(zillow_index) + time_trend | household_code | 0 | 
                       dma_code + panel_year, data = shares_DT)

# create recession (Dec07 through Jun09) dummy indicator
shares_DT[, recession := month_year >= mdy("12-01-2007") & 
            month_year <= mdy("06-01-2009")]

# time trend plus recession indicator (dummy) model
fit_timetrend_rec = felm(perc_share ~ log(income) + factor(unemployed) +
                           log(zillow_index) + time_trend + 
                           factor(recession) | household_code | 0 | 
                           dma_code + panel_year, data = shares_DT)

# month/year fixed effects model
fit_monthyear = felm(perc_share ~ log(income) + factor(unemployed) +
                       log(zillow_index) | household_code + month_year | 0 | 
                       dma_code + panel_year, data = shares_DT)

```

To save memory, we only kept the complete observations in `shares_DT`:

```{r}
# keep only complete cases
shares_DT = shares_DT[complete.cases(shares_DT)]
```

\newpage

Finally, we used the `stargazer` package to combine and display the results of the models:

```{r}
# stargazer for model comparison
stargazer(fit_base, fit_demo, fit_household, type = "text", 
          column.labels = c("Base", "Demographics", "Household FE"), 
          dep.var.labels.include = FALSE)
```

\newpage

```{r}
stargazer(fit_timetrend, fit_timetrend_rec, fit_monthyear, type = "text", 
          column.labels = c("Time Trend", "Time Trend w/Rec Dummy", 
                            "Month/Year FE"), dep.var.labels.include = FALSE)
```

\newpage

## Discussion of the results

Our model estimates reflected how percentage of private label share changes once we control for factors of recession as well as including the decrease of income and household values. Additionally, the women as head of the household seems to be positively correlated with the increase of private label purchases. As we controlled for household fixed effects, we were able to see a potential impact in private label purchases driven by household values. Unemployment did not seem to be affected in this scenario.

In reference to time trends, our model suggested that there was strong statistical evidence that economic downturns such as recessions affect private label demand. When controlling for recessions, the decrease in housing values only slightly affects private label purchase shares. 

This recession factor also has a credible causal interpretation, as it is a variable has constant impact across certain time frame for all households. We can infer from the estimates that income effect is much lower with fixed effects (within household) than without and the coefficient is consistent even when the time trend is applied. The recession dummy had an impact on the income and wealth for households, showing the causal effect. This is further supported by the R2 and adjusted R2 that don’t change much once the time trend is applied with fixed effects.

In particular, the largest observed change in residual error in our models (12.804 for base and 12.739 for demographics to 8.819 for household FE) was when we added the household fixed effects. This is likely due to the within estimation and eliminating for any cross group variation due to any unobserved variables (omitted variable bias).

We also considered the implied economic magnitudes of the income, wealth, and employment effects on private label shares. Using our last model (household and month/year FE), we estimated the effect of the following changes on private label shares:
* Reduce 50% income: increase 0.13% share of private label
* Reduce 50% wealth: increase 0.26% share of private label
* Become unemployed: increase 0.36% share of private label

Overall, private label demand seems to be sensitive to household income and wealth. This could support why many retailers such as Save-A-Lot Food Stores, a limited assortment grocery store chain that specifically targets low-income households, saw an increase in sales during economic downturns.
