---
title: "Step 1: Estimation and prediction of conditional average treatment effects"
author: "Jack Gang"
date: "3/15/2018"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
header-includes: \usepackage{color}
graphics: yes
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, comment = NA, message = FALSE,
                      fig.width = 10, fig.height = 10, fig.align = "center")
```

In this step, I used the 2015 data to estimate and validate several models to predict the heterogeneous treatment effects:

```{r, cache = TRUE}
# load libraries
library(bit64)
library(data.table)
library(glmnet)
library(causalTree)
library(ggplot2)
library(knitr)
library(corrplot)
library(broom)

# load data and add causal forest estimates
load("/classes/37105/main/Assignment-7/Customer-Development-2015.RData")
load("CATE-Causal-Forest.RData")
crm_DT[, tau_cforest := predict_DT_2015$tau_cforest]
```

First, I split the sample into a 50% training and 50% validation sample:

```{r, cache = TRUE}
# split data into training and validation samples
set.seed(2001)
crm_DT[, training_sample := rbinom(nrow(crm_DT), 1, 0.5)]
```

To make the code more readable, I renamed the `mailing_indicator` to make it clear that the randomized mailing is the treatment, W:

```{r, cache = TRUE}
# rename mailing_indicator
setnames(crm_DT, "mailing_indicator", "W")
```

\newpage

### Data pre-processing

As in the previous assignment, I removed highly correlated features from the data set. First I calculated a matrix of correlation coefficients among all inputs:

```{r}
# correlation matrix of inputs
cor_matrix = cor(crm_DT[, !c("customer_id", "W", "outcome_spend"), 
                        with = FALSE])
```

I then created a data table that contains the correlations for all variable pairs:

```{r}
# correlation data table
cor_matrix[upper.tri(cor_matrix, diag = TRUE)] = NA
cor_DT = data.table(row = rep(rownames(cor_matrix), ncol(cor_matrix)),
                    col = rep(colnames(cor_matrix), each = ncol(cor_matrix)),
                    cor = as.vector(cor_matrix))
cor_DT = cor_DT[is.na(cor) == FALSE]
remove(cor_matrix)
```

Finally, I found all correlations larger than 0.95 in absolute value. I inspected these correlations and eliminated one of the virtually redundant variables in each highly correlated pair from the data set:

```{r, cache = TRUE}
# find all correlations larger than 0.95 in absolute value
large_cor_DT = cor_DT[abs(cor) > 0.95]
kable(large_cor_DT, digits = 4)

# eliminate redundant variables in the row column
crm_DT = crm_DT[, !large_cor_DT$row, with = FALSE]
remove(cor_DT)

# acquisition_days_since, acquisition_months_since, and in_database_months are
# highly correlated, and the line above removes two out of the three and keeps
# acquisition_months_since
```

\newpage

### Randomization checks

I inspected the data to estimate the probability of a mailing (the propensity score). I also performed a logistic regression as a quick check to assess if the treatment was indeed randomized:

```{r}
# propensity score
propScore = mean(crm_DT$W)
paste("Probability of a mailing:", propScore)

# ensure catalog mailing is indeed randomized
fit = glm(W ~ ., family=binomial(), data=crm_DT)
results_DT = as.data.table(tidy(fit))
kable(results_DT[p.value < 0.01], digits = 5)
remove(fit)
remove(results_DT)
```

As seen in the table above, there are only four customer attributes that have better than 0.001 significance in predicting the catalog mailing indicator. Ouf ot the four, three seem directly related to outcome spend (`spend_period_1b`, `orders_d_1yr`, `spend_notz_1yr`) which we hypothesize can be attributed to the mailing indicator, and the last one, `customer_id`, has an estimate of zero.

### Estimation of heterogeneous treatment effects

I used the training sample to estimate the conditional average treatment effect on dollar spending due to catalog targeting. The following models were estimated: OLS, LASSO, and Causal forest:

```{r, cache = TRUE}
# create separate training and validation samples
training_DT = crm_DT[training_sample == 1, 
                     !c("customer_id", "training_sample", "tau_cforest"), 
                     with = FALSE]
validation_DT = crm_DT[training_sample == 0,
                       !c("customer_id", "training_sample", "tau_cforest"), 
                       with = FALSE]

# OLS
ols_fit = lm(outcome_spend ~ . + .:W - W, data = training_DT)

# LASSO
x = model.matrix(outcome_spend ~ 0 + . + .:W - W, data = training_DT)
y = training_DT$outcome_spend

lasso_fit = glmnet(x, y)
plot(lasso_fit, xvar = "lambda")

cv_lasso_fit = cv.glmnet(x, y)
cv_lasso_fit$lambda.min
cv_lasso_fit$lambda.1se

#coef(cv_lasso_fit, s = "lambda.min")
plot(cv_lasso_fit)

# causal forest - use estimates supplied by professor already in crm_DT

# save output objects for later use
load("ols_fit.RData")
load("lasso_fit.RData")
load("cv_lasso_fit.RData")
# save(ols_fit, file = "ols_fit.RData")
# save(lasso_fit, file = "lasso_fit.RData")
# save(cv_lasso_fit, file = "cv_lasso_fit.RData")
```

\newpage

### Predict treatment effects

To validate the models, I predicted the treatment effects and saved them to a file:

```{r, cache = TRUE}
# make separate validation_DT's with W set to 0 and 1 to calculate treatment effect
valW1_DT = cbind(validation_DT)
valW1_DT[, W := 1]
valW0_DT = cbind(validation_DT)
valW0_DT[, W := 0]

# predict tau for models
pred_y_OLS_W1 = predict(ols_fit, newdata = valW1_DT)
pred_y_OLS_W0 = predict(ols_fit, newdata = valW0_DT)
pred_y_OLS = pred_y_OLS_W1 - pred_y_OLS_W0

x_val_W1 = model.matrix(outcome_spend ~ 0 + . + .:W - W, data = valW1_DT)
x_val_W0 = model.matrix(outcome_spend ~ 0 + . + .:W - W, data = valW0_DT)
pred_y_LASSO_W1 = predict(cv_lasso_fit, newx = x_val_W1, s = "lambda.min")
pred_y_LASSO_W0 = predict(cv_lasso_fit, newx = x_val_W0, s = "lambda.min")
pred_y_LASSO = pred_y_LASSO_W1 - pred_y_LASSO_W0

predict_DT = crm_DT[training_sample == 0, !c("customer_id", "training_sample"), 
                    with = FALSE][, .(W, outcome_spend, tau_cforest)]
predict_DT[, tau_ols := pred_y_OLS]
predict_DT[, tau_lasso := pred_y_LASSO]

remove(x, y, training_DT, validation_DT, x_val_W1, x_val_W0, pred_y_OLS_W1,
       pred_y_OLS_W0, pred_y_LASSO_W1, pred_y_LASSO_W0, pred_y_OLS, pred_y_lasso,
       valW0_DT, valW1_DT)

# save model predictions to a file
# remove(predict_DT)
# remove(ols_fit, lasso_fit, cv_lasso_fit)
load("predict_DT.RData")
# save(predict_DT, file = "predict_DT.RData")
```






