---
title: "lfe: Linear Regression With Many Fixed Effects"
author: "Günter J. Hitsch"
date: "January 2018"
output: pdf_document
urlcolor: blue
graphics: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE,
                      fig.width = 4.5, fig.height = 3, fig.align = "right")
```




## Why use lfe?

The lfe package estimates linear regression models with a large number of fixed effects

In the standard `lm` function we can estimate fixed effects as dummy variables. For example, if `id` is the identifier of a store and we want to estimate the effect of `x` on `y` while controlling for store fixed effects, then we estimate the regression:

```{r, eval = FALSE}
fit = lm(y ~ x + factor(id), data = DT)
```

However, for a large number of fixed effects the `lm` function becomes inefficient and slow and eventually breaks down. As a rule of thumb, I would not use `lm` with more than about 500 fixed effects.

lfe is a great alternative to `lm`. It is designed to efficiently estimate regression models even with tens of thousands or more fixed effects.



## Documentation

I am not aware of any good tutorial but usage of the package is straightforward. The official documentation is here:

<https://cran.r-project.org/web/packages/lfe/lfe.pdf>



\newpage

## Usage example

Let's simulate a data set with price and quantity observations in a panel data set with many stores and weeks. Experiment with the settings for the number of stores and weeks and check how quickly the lfm package can obtain estimates!

```{r, results = "hide"}
library(data.table)
library(lfe)

# Settings: Number of stores and weeks
N_stores = 200
N_weeks  = 100

# Settings: Regression coefficients, fixed effects, and error term standard deviation
price_effect  = -2.5
store_effects = 30 + 4*round(sin(1:N_stores), 1)
week_effects  = round((0:(N_weeks-1))*(2/(N_weeks-1)), 1)
error_sd      = 1.5

# Total number of observations
N = N_stores*N_weeks 

set.seed(1776)
DT = data.table(store_id = as.integer(rep(1:N_stores, each = N_weeks)),
                week     = as.integer(rep(1:N_weeks, times = N_stores)),
                price    = round(runif(N, min = 4, max = 10), digits = 1)
               )

# Purchase quantity simulation
DT[, quantity := price_effect*price + store_effects[store_id] + week_effects[week]
                 + round(rnorm(N, sd = error_sd), 1)]
```

```{r}
head(DT)
```



\newpage

#### Estimation

Now estimate the regression model using the `felm` (fixed-effects linear model) function:

```{r}
fit = felm(quantity ~ price | store_id + week, data = DT)
```

Note the syntax to add both store and week fixed effects to the regression formula.

The regression output will not include (fortunately!) the fixed effect estimates:

```{r}
summary(fit)
```



\newpage

#### Obtain the fixed effects estimates

If you need to obtain the fixed effect estimates, use `getfe`. The results are stored in a data frame.

```{r}
FE = getfe(fit, se = TRUE)

head(FE)
```


#### Plot the estimated vs. the true fixed effects

True and estimated store fixed effects:

```{r}
library(ggplot2)
setDT(FE)          # Convert FE to a data.table

ggplot(FE[fe == "store_id"], aes(x = store_effects, y = effect)) +
   geom_point()
```

True and estimated week fixed effects:

```{r}
ggplot(FE[fe == "week"], aes(x = week_effects, y = effect)) +
   geom_point()
```



\newpage

#### Prediction

The lfe package does not contain a `predict` method that works with `felm`. Hence we create the function `predict.felm` below. The script `predict.felm.R` contains the same code, and you can source the script whenever you need to predict using the output from a lfe regression.

```{r}
predict.felm <- function(fit, newdata) {
   
   if (class(fit) != "felm") stop("'fit' is not a felm object")
   if (!("data.frame" %in% class(newdata))) stop("'newdata' must be a data.frame or data.table")

   setDT(newdata)
   uses_FEs = length(fit$fe) > 0

   # Predict output based on estimated coefficients, not inclucing fixed effects
   formula_string = as.character(fit$terms)
   if (uses_FEs) original_formula = paste("~ 0 +", formula_string[2])
   else original_formula = paste("~ ", formula_string[2])
   X = model.matrix(formula(original_formula), newdata)

   # Only retain columns in X if among the original inputs
   X = X[, rownames(fit$coefficients)]

   y = as.vector(X %*% fit$coefficients)

   # Add fixed effect values to prediction
   if (uses_FEs) {
      FE = as.data.table(getfe(fit))
      cols = c("fe", "idx")
      FE[, (cols) := lapply(.SD, as.character), .SDcols = cols]

      for (name in unique(FE$fe)) {
         fe_DT = newdata[, name, with = FALSE]
         fe_DT[, obs_no := .I]
         setnames(fe_DT, name, "idx")
         fe_DT[, idx := as.character(idx)]
         
         fe_DT = merge(fe_DT, FE[fe == name, .(idx, effect)], by = "idx")
         fe_DT = fe_DT[order(obs_no)]
         
         y = y + as.vector(fe_DT$effect)
      }
   }
   
   return(y)
}
```

\bigskip

Let's test it:

```{r}
predicted_quantity = predict.felm(fit, DT)
cor(DT$quantity, predicted_quantity)
```

If we increase prices by 3:

```{r, results = "hide"}
new_DT = copy(DT)
new_DT[, price := price + 3]
new_predicted_quantity = predict.felm(fit, new_DT)
```

```{r}
mean(predicted_quantity)
mean(new_predicted_quantity)
```


